---
title: "Introduction to the ASSIGN Package"
author:
- name: Ying Shen
  affiliation: 
  - The Section of Computational Biomedicine, Boston University School of Medicine, Boston, MA
  - Program in Bioinformatics, Boston University, Boston, MA
- name: David Jenkins
  affiliation: 
  - The Section of Computational Biomedicine, Boston University School of Medicine, Boston, MA
  - Program in Bioinformatics, Boston University, Boston, MA
- name: W. Evan Johnson
  affiliation: 
  - The Section of Computational Biomedicine, Boston University School of Medicine, Boston, MA
  - Program in Bioinformatics, Boston University, Boston, MA
  email: wej@bu.edu
package: ASSIGN
output: 
  BiocStyle::html_document2:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Primer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette provides an overview of the Bioconductor package
`r BiocStyle::Biocpkg("ASSIGN")` (Adaptive Signature Selection and InteGratioN) for
signature-based profiling of heterogeneous biological pathways.
`r BiocStyle::Biocpkg("ASSIGN")` is a computational tool used to evaluate the pathway
deregulation/activation status in individual patient samples.
`r BiocStyle::Biocpkg("ASSIGN")` employs a flexible Bayesian factor analysis approach that
adapts predetermined pathway signatures derived either from a literature search
or from perturbation experiments to create cell-/tissue-specific pathway
signatures. The deregulation/activation level of each context-specific pathway
is quantified to a score, which represents the extent to which a patient sample
matches thepathway deregulation/activation signature.

Some distinctive features of `r BiocStyle::Biocpkg("ASSIGN")` are:

1. __Multiple Pathway Profiling__: `r BiocStyle::Biocpkg("ASSIGN")` can profile multiple
pathway signatures simultaneously, accounting for 'cross-talk' between
interconnected pathway components.
2. __Context specificity in baseline gene expression__: Baseline gene expression
levels (i.e., the gene expression level under normal conditions) may vary widely
due to differences across tissue types, disease statuses, or measurement
platforms. `r BiocStyle::Biocpkg("ASSIGN")` can adaptively estimate background gene
expression levels across a set of samples.
3. __Context-specific signature estimation__: `r BiocStyle::Biocpkg("ASSIGN")` provides the
flexibility to use either an input gene list or magnitudes of signature genes as
prior information, allowing for adaptive refinement of pathway signatures in
specific cell or tissue types.
4. __Regularization of signature strength estimates__: `r BiocStyle::Biocpkg("ASSIGN")`
regularizes the signature strength coefficients using a Bayesian ridge
regression formulation by shrinking the strength of the irrelevant signature
genes toward zero. The parameter regularization constrains the pathway signature
to a small group of genes, making the results more biologically interpretable.

# How to use the ASSIGN package

## Example Data

In the following examples, we will illustrate how to run `r BiocStyle::Biocpkg("ASSIGN")`
using either the easy to use `assign.wrapper` function for simple
analysis or each individual `r BiocStyle::Biocpkg("ASSIGN")` step for more detailed
intermediate results.

For either analysis, we will first load `r BiocStyle::Biocpkg("ASSIGN")` and create a
temporary directory `'tempdir'` under the user's current working directory. All
output generated in this vignette will be saved in `'tempdir'`.

```{r setup, echo=TRUE}
library(ASSIGN)

dir.create("tempdir")
tempdir <- "tempdir"
```

Next, load the training data, test data, training labels, and test labels. The
training dataset is a G (number of genomic measurements) x N (number of samples
in pathway perturbation experiments) matrix, including five oncogenic pathways:
B-Catenin, E2F3, MYC, RAS, and SRC pathways in this example. The training data
labels denote the column indices of control and experimental samples for each
perturbation experiment. For example, we specify the column indices of the 10
RAS control samples to be 1:10, and column indices of 10 RAS activated samples
to be 39:48. The test dataset is a G (number of genomic measurements) x N
(number of patient samples) matrix. The test data labels denote the classes of
the N test samples. In our example, test samples 1-53 are adenocarcinoma and
samples 54-111 are squamous cell carcinoma. We specify `'Adeno'` and
`'Squamous'` in the vector of test data labels. Note that the test data labels
are optional. `r BiocStyle::Biocpkg("ASSIGN")` outputs additional validation plots to
evaluate classification accuracy when test data labels are provided.

```{r datasets-and-labels, eval=FALSE}
data(trainingData1)
data(testData1)
data(geneList1)
trainingLabel1 <- list(control = list(bcat=1:10, e2f3=1:10,
                                      myc=1:10, ras=1:10, src=1:10),
                       bcat = 11:19, e2f3 = 20:28, myc= 29:38,
                       ras = 39:48, src = 49:55)
testLabel1 <- rep(c("Adeno", "Squamous"), c(53,58))
```

## Run `r BiocStyle::Biocpkg("ASSIGN")` all-in-one using `assign.wrapper`

We developed an all-in-one `assign.wrapper` function to run
`r BiocStyle::Biocpkg("ASSIGN")` with one command. For most users, `assign.wrapper` will be
sufficient. The `assign.wrapper` function outputs the following files:

* __pathway_activity_testset.csv__: `r BiocStyle::Biocpkg("ASSIGN")` predicted
  pathway activity in test samples.
* __signature_heatmap_testset_prior.pdf__: heatmaps of the
  expression level of prior signature genes in training samples.
* __pathway_activity_scatterplot_testset.pdf__: scatterplot of
  pathway activity in test samples. The x-axis represents test samples ordered
  by pathway activity; the y-axis represents pathway activity.
* __output.rda__: The intermediate results of individual
  `r BiocStyle::Biocpkg("ASSIGN")` functions.
* __parameters.txt__: A log file containing the parameters used for
  this `r BiocStyle::Biocpkg("ASSIGN")` run.

If training data is provided, `assign.wrapper` also outputs the
following files:

* __pathway_activity_trainingset.csv__: `r BiocStyle::Biocpkg("ASSIGN")` predicted
  pathway activity in training samples.
* __signature_heatmap_trainingset.pdf__: heatmaps of the expression
  level of signature genes in training samples.
* __pathway_activity_scatterplot_trainingset.pdf__: scatterplot of
  pathway activity in training samples.
* __signature_gene_list_prior.csv__: the gene list and prior
  coefficients for the pathway signature.

When `Adaptive_S` is TRUE, `assign.wrapper` also outputs the following files:

* __signature_heatmap_testset_posterior.pdf__: heatmaps of the
  expression level of posterior signature genes in training samples.
* __posterior_delta.csv__: a csv file of the prior and posterior
  change in expression and probability of inclusion for each gene in each
  signature.
* __Signature_convergence.pdf__: A plot of the MCMC convergence.

Finally, if the `testLabel` argument is not NULL,
`assign.wrapper` also outputs the following files:

* __pathway_activity_boxplot_testset.pdf__: boxplot of pathway
  activity in every test class.

Here we illustrate how to run `assign.wrapper` function with three
examples. To start, create a temporary directory `'tempdir'` and load training
and test datasets. The individual parameters are described in detail in the
sections below and the `r BiocStyle::Biocpkg("ASSIGN")` reference manual.

### Example 1: Training data is available, but a gene list of pathway signature genes is not available:

```{r all-in-one-assign-wrapper-example1, eval=FALSE, results='hide'}
dir.create(file.path(tempdir,"wrapper_example1"))
assign.wrapper(trainingData=trainingData1, testData=testData1,
               trainingLabel=trainingLabel1, testLabel=testLabel1,
               geneList=NULL, n_sigGene=rep(200,5), adaptive_B=TRUE,
               adaptive_S=FALSE, mixture_beta=TRUE,
               outputDir=file.path(tempdir,"wrapper_example1"),
               iter=2000, burn_in=1000)
```

### Example 2: Training data is available, and a gene list of pathway signature genes is available:

```{r all-in-one-assign-wrapper-example2, eval=FALSE, results='hide'}
dir.create(file.path(tempdir,"wrapper_example2"))
assign.wrapper(trainingData=trainingData1, testData=testData1,
               trainingLabel=trainingLabel1, testLabel=NULL,
               geneList=geneList1, n_sigGene=NULL, adaptive_B=TRUE,
               adaptive_S=FALSE, mixture_beta=TRUE,
               outputDir=file.path(tempdir,"wrapper_example2"),
               iter=2000, burn_in=1000)
```

### Example 3: Training data is not available, but a gene list of pathway signature genes is available:

```{r all-in-one-assign-wrapper-example3, eval=FALSE, results='hide'}
dir.create(file.path(tempdir,"wrapper_example3"))
assign.wrapper(trainingData=NULL, testData=testData1,
               trainingLabel=NULL, testLabel=NULL,
               geneList=geneList1, n_sigGene=NULL, adaptive_B=TRUE,
               adaptive_S=TRUE, mixture_beta=TRUE,
               outputDir=file.path(tempdir,"wrapper_example3"),
               iter=2000, burn_in=1000)
```

## Run `r BiocStyle::Biocpkg("ASSIGN")` step-by-step

We developed a series of functions: `assign.preprocess`, `assign.mcmc`,
`assign.convergence`, `assign.summary`, `assign.cv.output`, and `assign.output`
that work in concert to produce detailed results.

### `assign.preprocess`

We first run the `assign.preprocess` function on the input datasets.
When the genomic measurements (e.g., gene expression profiles) of training
samples are provided, but predetermined pathway signature gene lists are not
provided, the `assign.preprocess` function utilizes a Bayesian
univariate regression module to select a gene set (usually 50-200 genes, but
this can be specified by the user) based on the absolute value of the regression
coefficient (fold change) and the posterior probability of the variable to be
selected (statistical significance). Since we have no predetermined gene lists
to provide, we leave the `geneList` option as default NULL. Here we
specify 200 signature genes for each of the five pathways.

```{r assign-preprocess-function1, eval=FALSE, results='hide'}
# training dataset is available;
# the gene list of pathway signature is NOT available
processed.data <- assign.preprocess(trainingData=trainingData1,
                                    testData=testData1,
                                    trainingLabel=trainingLabel1,
                                    geneList=NULL, n_sigGene=rep(200,5))
```

Alternatively, the users can have both the training data and the
curated/predetermined pathway signatures. Some genes in the curated pathway
signatures, although not significantly differentially expressed, need to be
included for the purpose of prediction. In this case, we specify the
`trainingData` and `geneList` parameters when both the training
dataset and predetermined signature gene list are available.

```{r assign-preprocess-function2, eval=FALSE, results='hide'}
# training dataset is available;
# the gene list of pathway signature is available
processed.data <- assign.preprocess(trainingData=trainingData1,
                                    testData=testData1,
                                    trainingLabel=trainingLabel1,
                                    geneList=geneList1)
```

In some cases, the expression profiles (training dataset) is unavailable. Only
the knowledge-based gene list or gene list from the joint knowledge of some
prior profiling experiments is available. In this case, we specify
`geneList` and leave the `trainingData` and
`trainingLabel` as default NULL.

```{r assign-preprocess-function3, eval=FALSE, results='hide'}
# training dataset is NOT available;
# the gene list of pathway signature is available
processed.data <- assign.preprocess(trainingData=NULL,
                                    testData=testData1,
                                    trainingLabel=NULL,
                                    geneList=geneList1)
```

The `assign.preprocess` function returns the processed training
dataset (`trainingData_sub`) and test dataset (`testData_sub`) as well as the
prior parameters for the background vector (`B_vector`), signature matrix
(`S_matrix`) and the probability signature matrix (`Pi_matrix`) and
differentially expressed gene lists of each pathway (`diffGeneList`). The
details of the `assign.preprocess` output are described in the `'value'`
section of the manual page of `assign.preprocess` function. The output data of
`assign.preprocess` function are used as the input data of the `assign.mcmc`
function.

### `assign.mcmc`

For the `assign.mcmc` function, Y, Bg, and X are specified as the
output of the `assign.preprocess` function. The `adaptive_B`
(adaptive background), `Adaptive_S` (adaptive signature) and
`mixture_beta` (regularization of signature strength) can be specified
TRUE or FALSE based on the analysis context. When training and test samples are
from the different cell or tissue types, we recommend the adaptive background
option to be TRUE. Notice that when the training dataset is not available, the
adaptive signature option must be set TRUE, meaning that the magnitude of the
signature should be estimated from the test dataset. The default `iter`
(iteration) is 2000. Particularly, when training datasets are unavailable, it is
better to specify the `X` option in the `assign.mcmc` using a
more informative X (specify up- or down- regulated genes) to initiate the model.

```{r assign-mcmc-function, eval=FALSE, results='hide'}
mcmc.chain <- assign.mcmc(Y=processed.data$testData_sub,
                          Bg = processed.data$B_vector,
                          X=processed.data$S_matrix,
                          Delta_prior_p = processed.data$Pi_matrix,
                          iter = 2000, adaptive_B=TRUE,
                          adaptive_S=FALSE, mixture_beta=TRUE)
```

The `assign.mcmc` function returns the MCMC chain recording default
2000 iterations for each parameter. The details of `assign.mcmc`
output are described in the `'value'` section of the manual page of
`assign.mcmc` function.

### `assign.convergence`

We can make a trace plot to check the convergence of the model parameters
through `assign.convergence`. The `burn_in` default is 0, so
that the trace plot starts from the first iteration. Additional iterations
can be specified if the MCMC chain does not converge in 2000 iterations.

```{r assign-convergence-function, eval=FALSE, results='hide'}
trace.plot <- assign.convergence(test=mcmc.chain, burn_in=0, iter=2000,
                                 parameter="B", whichGene=1,
                                 whichSample=NA, whichPath=NA)
```

The `assign.convergence` function returns a vector of the estimated values from
each Gibbs sampling iteration of the model parameter to be checked and a trace
plot of the parameter.

### `assign.summary`

We then apply the `assign.summary` function to compute the posterior
mean of each parameter. Typically we use the second half of the MCMC chain to
compute the posterior mean. We specify the default burn-in period to be the
first 1000 iteration and the default total iteration to be 2000. The 1000
burn-in iterations are discarded when we compute the posterior mean. The
`adaptive_B`, `Adaptive_S` and `mixture_beta` options
have to set the same as those in the `assign.mcmc` function.

```{r assign-summary-function, eval=FALSE, results='hide'}
mcmc.pos.mean <- assign.summary(test=mcmc.chain, burn_in=1000,
                                iter=2000, adaptive_B=TRUE,
                                adaptive_S=FALSE, mixture_beta=TRUE)
```

The `assign.summary` function returns the posterior mean of each
parameter. The details of the `assign.summary` output are described in
the `'value'` section of the manual page of `assign.summary` function.

### `assign.cv.output`

The `assign.cv.output` generates the cross-validation results in the
training samples. Output files from `assign.cv.output` are:

* __pathway_activity_trainingset.csv__: `r BiocStyle::Biocpkg("ASSIGN")` predicted
  pathway activity in training samples.
* __signature_heatmap_trainingset.pdf__: heatmaps of the expression
  level of signature genes in training samples.
* __pathway_activity_scatterplot_trainingset.pdf__: scatterplot of
  pathway activity in training samples.

```{r assign-cv-output-function, eval=FALSE, results='hide'}
# For cross-validation, Y in the assign.mcmc function
# should be specified as processed.data$trainingData_sub.
assign.cv.output(processed.data=processed.data,
                 mcmc.pos.mean.trainingData=mcmc.pos.mean,
                 trainingData=trainingData1,
                 trainingLabel=trainingLabel1, adaptive_B=FALSE,
                 adaptive_S=FALSE, mixture_beta=TRUE,
                 outputDir=tempdir)
```

### `assign.output`

The `assign.output` generates the prediction results in the test
samples. Output files from `assign.output` are:

* __pathway_activity_testset.csv__: `r BiocStyle::Biocpkg("ASSIGN")` predicted
  pathway activity in test samples.
* __signature_heatmap_testset_prior.pdf__: heatmaps of the
  expression level of prior signature genes in training samples.
* __signature_heatmap_testset_posterior.pdf__: heatmaps of the
  expression level of posterior signature genes in training samples. This plot
  is only generated when `Adaptive_S` is specified TRUE.
* __pathway_activity_scatterplot_testset.pdf__: scatterplot of
  pathway activity in test samples. The x-axis represents test samples ordered
  by pathway activity; the y-axis represents pathway activity.
* __pathway_activity_boxplot_testset.pdf__: boxplot of pathway
  activity in every test class. This plot is only generated only the
  `testLabel` argument is not NULL.

The user needs to specify the output directory in the `outputDir` option, when
running `assign.cv.output` and `assign.output`.

```{r assign-output-function, eval=FALSE, results='hide'}
assign.output(processed.data=processed.data,
              mcmc.pos.mean.testData=mcmc.pos.mean,
              trainingData=trainingData1, testData=testData1,
              trainingLabel=trainingLabel1,
              testLabel=testLabel1, geneList=NULL,
              adaptive_B=TRUE, adaptive_S=FALSE,
              mixture_beta=TRUE, outputDir=tempdir)
```

# Additional Features

## Anchor Gene Lists and Exclude Gene Lists

The `r BiocStyle::Biocpkg("ASSIGN")` package allows a signature to be adapted to fit other
biological contexts, reducing the contribution of specific genes in the
signature to better match heterogeneity observed in the test dataset.
Occasionally, adapting a signature may reduce the importance of key signature
genes. For example, if a signature is created by overexpressing an oncogenic
gene in a cell line, but during the adaptation step, `r BiocStyle::Biocpkg("ASSIGN")`
reduces the importance of that key gene, the quality of the
`r BiocStyle::Biocpkg("ASSIGN")` predictions may be reduced. Alternatively, if a gene in
the signature is associated with some other heterogeneity in the data, such as
smoking status, `r BiocStyle::Biocpkg("ASSIGN")` may adapt to differences in that gene,
rather than the actual desired signature activity predictions. To this end, we
have added the ability to provide a list of key genes to anchor in the
signature, and genes to exclude from the signature. `r BiocStyle::Biocpkg("ASSIGN")`
accomplishes this by setting the probability of inclusion into the signature to
one for anchor genes, and zero for exclude genes. The change in expression
values can still adapt, increasing or reducing the fold change associated with
each gene in the signature, but the anchor genes will always contribute to the
final signature, and the exclude genes will not.

```{r anchor-exclude-example, eval=FALSE, results='hide'}
dir.create(file.path(tempdir, "anchor_exclude_example"))

anchorList = list(bcat="224321_at",
                  e2f3="202589_at",
                  myc="221891_x_at",
                  ras="201820_at",
                  src="224567_x_at")
excludeList = list(bcat="1555340_x_at",
                   e2f3="1555340_x_at",
                   myc="1555340_x_at",
                   ras="204748_at",
                   src="1555339_at")

assign.wrapper(trainingData=trainingData1, testData=testData1,
               trainingLabel=trainingLabel1, testLabel=NULL,
               geneList=geneList1, n_sigGene=NULL, adaptive_B=TRUE,
               adaptive_S=TRUE, mixture_beta=TRUE,
               outputDir=file.path(tempdir, "anchor_exclude_example"),
               anchorGenes=anchorList, excludeGenes=excludeList,
               iter=2000, burn_in=1000)
```

## Fraction of Upregulated Genes

By default, `r BiocStyle::Biocpkg("ASSIGN")` Bayesian gene selection chooses the signature
genes with an equal fraction of genes that increase with pathway activity and
genes that decrease with pathway activity. Use the pctUp parameter to modify
this fraction. Set pctUP to NULL to select the most significant genes,
regardless of direction.

# GFRN Optimization Procedure

When running `r BiocStyle::Biocpkg("ASSIGN")`, the number of genes in the gene list can
affect the predictions that `r BiocStyle::Biocpkg("ASSIGN")` produces, but it is not always
clear how long the gene list should be. Included within `r BiocStyle::Biocpkg("ASSIGN")` is
the optimization procedure used in the publication [Activity of distinct growth factor receptor network components in breast tumors uncovers two biologically relevant subtypes](http://dx.doi.org/10.1186/s13073-017-0429-x).
The function allows you to optimize the gene list lengths for the pathways
included in the paper using your own correlation data and gene list lengths.
This function runs `r BiocStyle::Biocpkg("ASSIGN")` pathway prediction on various gene list
lengths to find the optimum gene list length for the GFRN pathways by
correlating the `r BiocStyle::Biocpkg("ASSIGN")` predictions to a matrix of correlation
data that you provide. This function takes a long time to run because you are
running `r BiocStyle::Biocpkg("ASSIGN")` many times on many pathways, so I recommend
parallelizing by pathway or running the `r BiocStyle::Biocpkg("ASSIGN")` predictions first
(long and parallelizable) and then running the correlation step (quick)
separately.

## Example Optimization Procedure

The following example optimizes the pathway length for the AKT pathway based on
correlating `r BiocStyle::Biocpkg("ASSIGN")` predictions with proteomics data. First, read
in the test data that you want to predict using `r BiocStyle::Biocpkg("ASSIGN")` and the
data (e.g. proteomics data) that will be used for correlation:

```{r gfrn-optimization-dl, eval=FALSE, results='hide'}
dir.create(file.path(tempdir, "optimization_example"))
setwd(file.path(tempdir, "optimization_example"))

testData <- read.table("https://www.dropbox.com/s/6iuxcr2f0fjbirv/icbp_Rsubread_tpmlog.txt?dl=1",
                       sep='\t', row.names=1, header=1)

corData <- read.table("https://www.dropbox.com/s/nm3m38ca6v4lu10/proteomics.txt?dl=1",
                      sep='\t', row.names=1, header=1)
```

Next, create a list of data used for correlation. The list should contain a
vector of column names from the correlation data for each of the pathways
that are being optimized. The gene list length that has the largest average
correlation for the columns in the correlation list will be the optimized gene
list.

```{r gfrn-optimization-cor, eval=FALSE, results='hide'}
#this is a list of pathways and columns in the correlation data that will
#be used for correlation
corList <- list(akt=c("Akt","PDK1","PDK1p241"))
```

Finally, run the ComBat batch correction procedure and run the `optimizeGFRN`
function:

```{r gfrn-optimization-optimize, eval=FALSE, results='hide'}
#run the batch correction procedure between the test and training data
combat.data <- ComBat.step2(testData, pcaPlots = TRUE)

#run the default optimization procedure
optimization_results <- optimizeGFRN(combat.data, corData,
                                     corList, run="akt")
```

`r BiocStyle::Biocpkg("ASSIGN")` will output the results for each gene list length in the
current working directory. The `optimizeGFRN` function returns a list of
optimized gene lists which can be used on other datasets and correlation
results. Additional options and documentation is available in the
`optimizeGFRN` function documentation.

# Citing `r BiocStyle::Biocpkg("ASSIGN")`

If you use `r BiocStyle::Biocpkg("ASSIGN")` in your publication, please cite:

Shen, Y. et al. ASSIGN: context-specific genomic profiling of multiple
heterogeneous biological pathways. Bioinformatics 31 (11), 1745-1753 (2015).
[doi:10.1093/bioinformatics/btv031](http://dx.doi.org/10.1093/bioinformatics/btv031)

# Conclusion

Please see the `r BiocStyle::Biocpkg("ASSIGN")` reference manual for full descriptions of
functions and the various options they support.

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```